---
title: "Mapping Duiker OS"
author: "TaniaBarychka"
date: "20 March 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

Tutorial: https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

```{r echo=FALSE,warning=FALSE}

#install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", 
#"ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata"))

library("ggplot2")
theme_set(theme_bw())
library("sf")

library("rnaturalearth")
library("rnaturalearthdata")
library(rgeos)
```


The package rnaturalearth provides a map of countries of the entire world. Use ne_countries to pull country data and choose the scale (rnaturalearthhires is necessary for scale = "large"). The function can return sp classes (default) or directly sf classes, as defined in the argument returnclass:

```{r echo=FALSE,warning=FALSE}



world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)

## [1] "sf"  
## [1] "data.frame"

```


First, let us start with creating a base map of the world using ggplot2. This base map will then be extended with different map elements, as well as zoomed in to an area of interest. We can check that the world map was properly retrieved and converted into an sf object, and plot it with ggplot2:


```{r echo=FALSE,warning=FALSE}
ggplot(data = world) +
    geom_sf()

```


A title and a subtitle can be added to the map using the function ggtitle, passing any valid character string (e.g. with quotation marks) as arguments. Axis names are absent by default on a map, but can be changed to something more suitable (e.g. “Longitude” and “Latitude”), depending on the map:

```{r echo=FALSE,warning=FALSE}
ggplot(data = world) +
    geom_sf() +
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("World map", subtitle = paste0("(", length(unique(world$name)), " countries)"))

```


In many ways, sf geometries are no different than regular geometries, and can be displayed with the same level of control on their attributes. Here is an example with the polygons of the countries filled with a green color (argument fill), using black for the outline of the countries (argument color):

```{r echo=FALSE}
ggplot(data = world) + 
    geom_sf(color = "black", fill = "lightgreen")
ggplot(data = world) +
    geom_sf(aes(fill = pop_est)) +
    scale_fill_viridis_c(option = "plasma", trans = "sqrt")


```


The function coord_sf allows to deal with the coordinate system, which includes both projection and extent of the map. By default, the map will use the coordinate system of the first layer that defines one (i.e. scanned in the order provided), or if none, fall back on WGS84 (latitude/longitude, the reference system used in GPS). Using the argument crs, it is possible to override this setting, and project on the fly to any projection. This can be achieved using any valid PROJ4 string (here, the European-centric ETRS89 Lambert Azimuthal Equal-Area projection):


```{r echo=FALSE,warning=FALSE}

ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs ")
#takes a while to load
```


Spatial Reference System Identifier (SRID) or an European Petroleum Survey Group (EPSG) code are available for the projection of interest, they can be used directly instead of the full PROJ4 string. The two following calls are equivalent for the ETRS89 Lambert Azimuthal Equal-Area projection, which is EPSG code 3035:

```{r echo=FALSE, eval=FALSE, warning=FALSE}

ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = "+init=epsg:3035")

ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = st_crs(3035))

```


The extent of the map can also be set in coord_sf, in practice allowing to “zoom” in the area of interest, provided by limits on the x-axis (xlim), and on the y-axis (ylim). Note that the limits are automatically expanded by a fraction to ensure that data and axes don’t overlap; it can also be turned off to exactly match the limits provided with expand = FALSE:


```{r echo=FALSE,warning=FALSE}
ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE)

```


Several packages are available to create a scale bar on a map (e.g. prettymapr, vcd, ggsn, or legendMap). We introduce here the package ggspatial, which provides easy-to-use functions…

scale_bar that allows to add simultaneously the north symbol and a scale bar into the ggplot map. Five arguments need to be set manually: lon, lat, distance_lon, distance_lat, and distance_legend. The location of the scale bar has to be specified in longitude/latitude in the lon and lat arguments. The shaded distance inside the scale bar is controlled by the distance_lon argument. while its width is determined by distance_lat. Additionally, it is possible to change the font size for the legend of the scale bar (argument legend_size, which defaults to 3). The North arrow behind the “N” north symbol can also be adjusted for its length (arrow_length), its distance to the scale (arrow_distance), or the size the N north symbol itself (arrow_north_size, which defaults to 6). Note that all distances (distance_lon, distance_lat, distance_legend, arrow_length, arrow_distance) are set to "km" by default in distance_unit; they can also be set to nautical miles with “nm”, or miles with “mi”.

```{r echo=FALSE,warning=FALSE}

library("ggspatial")
ggplot(data = world) +
    geom_sf() +
    annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
        style = north_arrow_fancy_orienteering) +
    coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97))

## Scale on map varies by more than 10%, scale bar may be inaccurate

```


Note the warning of the inaccurate scale bar: since the map use unprojected data in longitude/latitude (WGS84) on an equidistant cylindrical projection (all meridians being parallel), length in (kilo)meters on the map directly depends mathematically on the degree of latitude. Plots of small regions or projected data will often allow for more accurate scale bars.

The world data set already contains country names and the coordinates of the centroid of each country (among more information). We can use this information to plot country names, using world as a regular data.frame in ggplot2. The function geom_text can be used to add a layer of text to a map using geographic coordinates. The function requires the data needed to enter the country names, which is the same data as the world map. Again, we have a very flexible control to adjust the text at will on many aspects:

- The size (argument size);
- The alignment, which is centered by default on the coordinates provided. The text can be adjusted horizontally or vertically using the arguments hjust and vjust, which can either be a number between 0 (right/bottom) and 1 (top/left) or a character (“left”, “middle”, “right”, “bottom”, “center”, “top”). The text can also be offset horizontally or vertically with the argument nudge_x and nudge_y;
- The font of the text, for instance its color (argument color) or the type of font (fontface);
- The overlap of labels, using the argument check_overlap, which removes overlapping text. Alternatively, when there is a lot of overlapping labels, the package ggrepel provides a geom_text_repel function that moves label around so that they do not overlap.
- For the text labels, we are defining the centroid of the counties with st_centroid, from the package sf. Then we combined the coordinates with the centroid, in the geometry of the spatial data frame. The package sf is necessary for the command st_centroid.
Additionally, the annotate function can be used to add a single character string at a specific location, as demonstrated here to add the Gulf of Mexico:

```{r echo=FALSE,warning=FALSE}
library("sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
geom_sf() +
geom_text(data= world_points,aes(x=X, y=Y, label=name),
    color = "darkblue", fontface = "bold", check_overlap = FALSE) +
annotate(geom = "text", x = -90, y = 26, label = "Gulf of Mexico", 
    fontface = "italic", color = "grey22", size = 6) +
coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE)
```


Now to make the final touches, the theme of the map can be edited to make it more appealing. We suggested the use of theme_bw for a standard theme, but there are many other themes that can be selected from (see for instance ?ggtheme in ggplot2, or the package ggthemes which provide several useful themes). Moreover, specific theme elements can be tweaked to get to the final outcome:

- Position of the legend: Although not used in this example, the argument legend.position allows to automatically place the legend at a specific location (e.g. "topright", "bottomleft", etc.);
- Grid lines (graticules) on the map: by using panel.grid.major and panel.grid.minor, grid lines can be adjusted. Here we set them to a gray color and dashed line type to clearly distinguish them from country borders lines;
- Map background: the argument panel.background can be used to color the background, which is the ocean essentially, with a light blue;
- Many more elements of a theme can be adjusted, which would be too long to cover here. We refer the reader to the documentation for the function theme.

```{r echo=FALSE,warning=FALSE}

ggplot(data = world) + geom_sf(fill="antiquewhite") + geom_text(data= world_points,aes(x=X, y=Y, label=name), color = "darkblue", fontface = "bold", check_overlap = FALSE) + annotate(geom = "text", x = -90, y = 26, label = "Gulf of Mexico", fontface = "italic", color = "grey22", size = 6) + annotation_scale(location = "bl", width_hint = 0.5) + annotation_north_arrow(location = "bl", which_north = "true", pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"), style = north_arrow_fancy_orienteering) + coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE) + xlab("Longitude") + ylab("Latitude") + ggtitle("Map of the Gulf of Mexico and the Caribbean Sea") + theme(panel.grid.major = element_line(color = gray(.5), linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "aliceblue"))

```
